---
title: "ACF functions in `itsadug`"
author: "Jacolien van Rij"
date: "version `r format.Date(Sys.Date(), '%B %d, %Y')`"
bibliography: bibliography.bib
output:   
  rmarkdown::html_document:
    fig_caption: yes
    toc: true
    theme: readable
    highlight: default
vignette: >
  %\VignetteIndexEntry{ACF functions in `itsadug`}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
<style>
td      {padding:5pt;}
p       {margin-top: 2em;}
</style>

The functions listed in this vignette apply to linear regression models, linear mixed models, and GAMMs (i.e., the functions are tested with `lm`, `glm`, `lmer`, `glmer`, `gam`, and `bam` models).

```{r}
suppressMessages(library(itsadug))
info('version')
```

## Example GAMM model
The code below was used to fit a GAMM model `m1` to the data set `simdat` from the package `itsadug`. The data set `simdat` is simulated time series data with arbitrary predictors. 

```{r, results='hide'}
data(simdat)

# For illustration purposes, we build a GAMM model
# with a nonlinear interaction, two groups, and
# random wiggly smooths for Subjects:
m1 <- bam(Y ~ Group + te(Time, Trial, by=Group)
  + s(Time, Subject, bs='fs', m=1),
  data=simdat)
```

## Visualizing the ACF of model residuals with `acf_resid`

The function `acf_resid` is a wrapper around the functions [`acf_plot`](#acfplot) and [`acf_n_plots`](#acfnplots). It allows for different ways of checking the ACF.

### 1. standard ACF

The default `acf` function R plots the autocorrelation function of the residuals as if the residuals are a single time series:
```{r, fig.width=4, fig.height=4}
acf(resid(m1))
```

Alternatively, the function `acf_resid` of the package `itsadug` could be used. This function offers different possibilities, as listed below:

```{r, fig.width=4, fig.height=4}
acf_resid(m1)
```


### 2. Average ACF over different time series

Individual time series could be provided as a named list, or as a vector with model predictors.

```{r, eval=FALSE}
# Option A: include named list
acf_resid(m1, split_pred=list(simdat$Subject,simdat$Trial))

# Option B: include model predictors
# This method only works for predictors that are included in the model.
acf_resid(m1, split_pred=c("Subject","Trial"))
```
```{r, fig.width=4, fig.height=4, echo=FALSE}
acf_resid(m1, split_pred=c("Subject","Trial"))
```

By default, function `acf_resid` calls `acf_plot` to calculate the averages of the time series. However, different measures can be provided with the argument `fun` in `acf_plots`:

```{r, fig.width=4, fig.height=4, results='hold'}
# Minimum ACF per lag:
acf_resid(m1, split_pred=c("Subject","Trial"), fun=min)
# Maximum ACF per lag:
acf_resid(m1, split_pred=c("Subject","Trial"), fun=max)
```

The function optionally returns the acf values, which can be used for generate more advanced ACF plots:

```{r, fig.width=4, fig.height=4}

# Median ACF per lag:
acf_resid(m1, split_pred="Subject", fun=median, lwd=3,
          main="Distribution of ACF")
# Calculate 25% and 75% quantiles:
acf1 <- acf_resid(m1, split_pred="Subject", 
    fun=function(x){quantile(x, .25)}, plot=FALSE)
acf2 <- acf_resid(m1, split_pred="Subject", 
    fun=function(x){quantile(x, .75)}, plot=FALSE)
# Plot these as error bars in different colors:
len <- length(acf1)-1
fill_area(x=0:len, y=acf2, from=acf1, col=alpha(1))
addInterval(pos=0:len, acf1, acf2, horiz=FALSE, col=alpha(1))
# add legend:
legend('topright',
    fill=alpha(1),
    border=alpha(1),
    legend='25-75%',
    bty='n')
```

### 3. N different ACF plots

The function `acf_resid` makes use of the function `acf_n_plots` to plot individual time series when the argument `n` is specified. 

##### Quantiles

By default `n` time series are plotted that represent $N$ quantiles (with respect to the value of lag 1).

```{r, fig.width=8, fig.height=6}
acf_resid(m1, split_pred=c("Subject","Trial"), n=6)
```

Optionally, the function outputs the quantiles:
```{r, fig.width=8, fig.height=6}
out <- acf_resid(m1, split_pred=c("Subject","Trial"), n=6, plot=FALSE)
# print the head of the elements in the first quantile:
head(out[[1]][['elements']])
# print the quantile:
out[[1]][['quantile']]
```


##### Random events

When `random=TRUE`, $N$ random events are being plotted:
```{r, fig.width=8, fig.height=6}
acf_resid(m1, split_pred=c("Subject","Trial"), n=6, random=TRUE)
```


##### Selection

With the argument `cond` (see `help(acf_pn_plots)`) specific events could be plotted:

```{r, fig.width=8, fig.height=6}
simdat$Event <- with(simdat, interaction(Subject, Trial))
acf_resid(m1, split_pred=list(Event=simdat$Event), n=6, 
    cond=list(Event=c('c05.-10', 'c11.-10', 'a05.-9', 'a09.-9', 'a13.-9', 'c02.-9')))
```


### 4. Output

The function `acf_resid` optionally gives back information about individual timeseries:

```{r}
# default output is the acf values:
(out <- acf_resid(m1, split_pred=c("Subject","Trial"), plot=FALSE))

# Alternatively, more information could be retrieved:
out <- acf_resid(m1, split_pred=c("Subject","Trial"), plot=FALSE, return_all=TRUE)
# out is a list of info:
names(out)

# 1. acf gives the acf values:
out[['acf']]

# 2. acftable provides the individual acf's in wide table format:
head(out[['acftable']], 3)
dim(out[['acftable']])

# 3. dataframe prvides a data frame with the acf, n, and ci information
# in long table format:
head(out[['dataframe']])

# 4. n provides the number of data points underlying each ACF:
head(out[['n']])
     
# 5. series and FUN provide info on input and function:
out[['series']]
out[['FUN']]
```

The data frames are useful for plotting the ACFs using other packages. This is an example from the vignette of the article of @BatesEtal :

```{r, eval=FALSE}
# Plot individual participants with the package lattice:
library(lattice)
out <- acf_resid(m1, split_pred=c("Subject"), plot=FALSE, return_all=TRUE)$dataframe
civec = out[out$lag==0,]$ci
xyplot(acf ~ lag | event, type = "h", data = out, col.line = "black", 
            panel = function(...) {
                panel.abline(h = civec[panel.number()], col.line = "grey")
                panel.abline(h = -civec[panel.number()], col.line = "grey")
                panel.abline(h = 0, col.line = "black")
                panel.xyplot(...)
            }, 
            strip = strip.custom(bg = "grey90"), 
            par.strip.text = list(cex = 0.8),
            xlab="lag", ylab="autocorrelation")
```

### 5. Correcting for AR1 $\rho$

When an AR1 model is included in a `gam` or `bam` model, the function `acf_resid` autmatically corrects for it:

```{r, fig.width=4, fig.height=4, results='hold'}
# genetare AR start column:
simdat <- start_event(simdat, column="Time", event="Event")
head(simdat)

# run GAMM with AR1 model:
m1 <- bam(Y ~ Group + te(Time, Trial, by=Group)
  + s(Time, Subject, bs='fs', m=1),
  data=simdat, rho=.65, AR.start=simdat$start.event)

# plot normal acf, without correction for rho:
acf(resid(m1))
# plot normal acf with acf_plot:
acf_resid(m1)
# plot normal acf with acf_plot:
acf_plot(resid(m1), split_by=list(simdat$Subject))
# plot corrected acf plot with acf_plot:
acf_plot(resid_gam(m1, incl_na=TRUE), split_by=list(simdat$Subject))
```



## The use of `acf_plot` <a id="acfplot"></a>

The function `acf_plot` is used for generating the ACF for individual time series, and may plot the averaged ACF. In contrast with `acf_resid` the input needs to be a vector, and the grouping predictors are provided to the argument `split_by` as a list with vectors.

```{r, fig.width=4, fig.height=4, results='hold'}
acf_plot(resid_gam(m1))
acf_plot(resid_gam(m1, incl_na=TRUE), split_by=list(simdat$Subject))
```


## The use of `acf_n_plots` <a id="acfnplots"></a>

The function `acf_n_plots` is used for generating $N$ ACF plots of individual time series. In contrast with `acf_resid` the input needs to be a vector, and the grouping predictors are provided to the argument `split_by` as a list with vectors.

```{r, fig.width=8, fig.height=6, results='hold'}
acf_n_plots(resid_gam(m1, incl_na=TRUE), split_by=list(simdat$Subject), n=6, random=TRUE)
```

# References


